hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-02-27 inst/manuscript/paper_pdf/paper.Rmd
add Morgan's edits to manuscript and re-render with edits

Morgan Sparks^1,4^, Eli Felts^2^, Allison Swartz^3^, Bryan Maitland^1^

@@

Understanding the timing of key life history events is necessary for managing and conserving populations. Historically, models to predict hatch and emergence timing for fishes were difficult to employ in wild settings because average incubation temperature was needed as the primary parameter in predictive models. However, recent improvements to these techniques reworked models such that they could be applied in wild environments as long as users had data for when adult fish spawned and a record of average daily temperature over the course of development. Despite these improvements, their application remains limited due to few parameterizations for varying species, being largely limited to salmonids. Here we present **hatchR**, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses. **hatchR** allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package. **hatchR** is also distributed in two forms—an open source R package for maximum customization, as well as an HTML graphical-user-interface web application for individuals not familiar with scripting languages. To demonstrate potential uses, we present two case studies as likely applications for this software. **hatchR** promises to open many exciting avenues in research and management of fishes during their early life history

@@

The resulting effective value approach has now been widely applied in Salmonids for which parameterizations from aquaculture were readily available—for example Pacific Salmon (*Oncorhynchus spp.*) models developed by @beacham1990 have been applied to various species and populations [@adelfio2024; @adelfio2019; @kaylor2021] while models developed for Bull Trout (*Salvelinus confluentus*) by @mcphail1979 were extended by @austin2019. Despite growing popularity, applications have been largely limited within Salmonids, presumably because parameterizations for such models already existed due to their wide use in aquaculture and their general popularity as sport and commercial fish.

To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, **hatchR**. Specifically, **hatchR** allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own or published temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.

To widen the user application of these methods, we distribute two user-interfaces for **hatchR**. The first is a R package (@r:ther) distributed via CRAN that allows users the most customizable application for these methods. The R package is especially powerful as it allows users to automate their analyses over multiple variables such as phenology type, multiple spawn dates, or different habitats with varying thermal regimes. These variable approaches are outlined in the package documentation on **hatchR**'s website. Alternatively, we also distribute a Shiny application (@chang2024) in the form of an HTML-based web tool to interact with many of **hatchR**'s functions in a graphical-user-interface. The Shiny form trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of **hatchR**'s R package. Below, we present the basic overview of the software and multiple case studies of how it may be applied.

@@

**hatchR** is meant to primarily be a tool for predicting phenology. In this sense, we limit functionality to these applications and provide minimal data checking and plotting help. This decision is in part driven by the diversity of data types that users may import and the difficulty in addressing all those data types with respect to various data checks. In other words, we expect users to know their data better than we do and to check it accordingly. We do provide two basic data check functions discussed in the Checking Data section. Similarly, we provide limited functionality for plotting results, but provide examples of how to build custom visualizations from output, specifically in R. For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in programs of the user's choice. Finally, we provide brief summaries of general applications of **hatchR** below, but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables.

```{r, echo = FALSE, out.height = '75%', fig.align = 'center', fig.cap="hatchR workflow. Data inputs are represented by folders, processes by filled circles, functions as plain text rectangles, and decisions as italicized text."}
knitr::include_graphics("./workflow.png")
```

@@

Effective value models were created by @sparks2019 to implement developmental models in wild environments for Sockeye Salmon (*O. nerka*). The need for their development arose because historic models, specifically those in @beacham1990, only considered the average incubation temperature during development and, for wild fishes, average incubation temperature was impossible to estimate because it was unknown when fish hatched even if adult spawn timing was known. To address this, @sparks2019 used the reciprocal of the formulation of model 2 from @beacham1990 and assigned an effective value for every day of development using the daily average temperature.

@@

Water temperature datasets collected for wild environments are often either 1.) already summarized by day (*i.e.*, mean daily temperature) or, 2.) in a raw format from something like a HOBO TidbiT where readings are taken multiple times per day, which can be summarized into a mean daily temperatures. Alternatively, new statistical models like that of @siegel2023 could be similarly implemented.

@@

**hatchR** assumes you've checked for missing records or errors in your data as it *will function with gaps*, so it's important to go through the data checks discussed below, as well as your own validity checks. **hatchR** can use values down to freezing (e.g, 0 °C), which returns extremely small effective values, and time to hatch or emerge may be \> 1 year. In these cases, we suggest users consider how much of that data type is reasonable with their data.

@@

We demonstrate the utility of the summary and check functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year-long dataset (`year_sim`, first 6 rows shown) with temperature readings every thirty minutes.

@@

head(year_sim)

@@

First, we recommend checking imported data for any outliers or strange inputs using `plot_check_temp()`

```{r, echo = TRUE, fig.cap="Output of hatchR function `plot_check_temp()`, which is used as a visual data check. Users can set custom thresholds for minimum and maximum temperatures (dashed lines)."}
plot_check_temp(data = year_sim,
                dates = date,
                temperature = temp,
                temp_min = 0, # temp_min and max lines are 
                              # user customizable
                temp_max = 25)
```

@@

Users can either use Salmonid models from `model_table` included in the package and the Shiny app. As discussed, these models are included because parameterizations already existed in the literature. However, these parameterizations are limited to Pacific Salmon and Bull Trout (though see @quinn2018 pg. 183 for numerous other Salmonids) . To widen the application of the effective value approach, we include a `fit_model()` function, which is species agnostic (as long as development generally follows the power law). Additionally, we limit available models in `model_table` because they are well vetted models, with experimental ranges spanning 2-17 °C. Custom models we demonstrate later with `fit_model()` have much narrower parameterization ranges and, as such, we choose to keep those models out of `model_table` so as to require users to think hard about whether the custom models they are parameterizing fit the data over which they are predicting.

The function `fit_model()` uses data in which average incubation temperature (°C) and days to phenological event (as two vectors) are the inputs. The function estimates parameter coefficients for *log~e~a* and *b* using `stats::nls()`

@@

Below, we demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package. We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for fishes beyond the scope of the original effective value approach. These parameterizations are for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*) from @webster1948, Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005. We provide parameterization for just Smallmouth Bass here for concision, but the full code set for all species is available at <https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper.Rmd>.

@@

```{r, echo = TRUE}
###  make temp regime
set.seed(123)
# create random temps and corresponding dates
temps_sim <- sort(rnorm(n =30, mean = 16, sd = 1), decreasing = FALSE)
dates_sim <-  seq(from = ymd("2000-07-01"),
             to = ymd("2000-07-31"), length.out = 30)

@@

# change names so they aren't the same as the vector objects
colnames(data_sim) <- c("temp_sim", "date_sim")

@@

# change F to C and hours to days
smallmouth <- smallmouth |> 
  mutate(days = ceiling(hours/24),
         temp_C = (temp_F -32) * (5/9))

smb_mod <- fit_model(temp = smallmouth$temp_C,
                     days = smallmouth$days,
                     species = "smb",
                     development_type = "hatch")

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted days to hatch for three warmwater species with custom built parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days."}
### predict_phenology

@@

To illustrate model selection and phenology prediction we will recreate a small portion of the analysis done by @sparks2019 using the `woody_island` dataset included in this package . We will predict both hatch and emergence timing, so we will obtain a model expression for each using `model_select()`.

@@

And then look inside the returned object to see days to hatch and development period:

```{r, echo = TRUE}
WI_hatch$days2done
WI_hatch$dev.period
```

@@

`WI_hatch$days2done` outputs the predicted days to develop.

`WI_hatch$dev.period` is a 1x2 dataframe with the dates corresponding to when your fish's parent spawned (which you input with `predict_phenology(spawn.date = ...)`) and the date when the fish is predicted todevelop.

`WI_hatch$ef.vals` is a vector of each day's effective value as evaluated using whatever model is chosen.

`WI_hatch$ef.tibble` is a *n* x 4 tibble (*n* = number of days to hatch or emerge) and the columns are the date, each day's temperature and effective value, and the cumulative sum of the effective values. The `ef.tibble` object is meant to serve as the basis for users to make custom figures for their data beyond the functionality we discuss below.

@@

`{r"} plot_phenology(WI_hatch)`

@@

We demonstrate this first case study using the graphical user interface portion of the **hatchR** ecosystem found at <https://elifelts.shinyapps.io/hatchR/>. Users will first upload their data with the `Import Data` window, which requires them to select their file on their personal computer, provide the program with the columns corresponding for dates and temperatures, and then provide the format in which dates are coded (*e.g.*, year-month-day or day-month-year). Data used in this case study are from the `crooked_river` data set from R and can either be written out from the package or accessed at <https://github.com/bmait101/hatchR/blob/master/extdata/crooked_river.csv/>. Once data is uploaded the program automatically plots the user's data using `plot_check_temp()` in the background and provides them the outputted graphical check. After uploading and checking data, the user switches to the `Model Phenology` window. In this circumstance, we use the preloaded parameterization for bull trout from @austin2019 with the `Existing` button for model selection, which the user selects with the various drop down options in the menu. After the model is selected, the user can choose multiple spawn dates from the interactive calendar provided. We show results for spawning for September 30th as indicated in the example above. Once dates are chosen, a table entry for each spawn date is outputted in the `Phenology Summaries` tab and corresponding plot with data from each spawn date in the `Timeline Plot` tab. Output from predicting phenology and the resulting figure are downloadable from their respective tabs. The process is demonstrated in full in Figure XXX, but the interface is described more completely on **hatchR**'s Shiny website.

@@

In this example we expect the last fish out of the gravel well before the June 1st date and the manager could allow grazing in this area without worrying about direct mechanical disturbance to fish developing in the gravel.

@@

For the second example, we will again use bull trout, but demonstrate a much more complex application for the purpose of showing the full flexibility of the programmatic application of **hatchR**. In this scenario we will use data from @isaak2018 (included `idaho` dataset), which includes temperature data from 226 sites across the major upper Columbia River headwater watersheds in Idaho. For this approach we winnow putative bull trout spawning sites by filtering for sites with mean August temperature \</= 13 °C in accordance with thresholds from @isaak2015. For the resulting 139 sites we will demonstrate predicting hatch timing in these putative Bull Trout spawning habitats.

@@

We need to setup our models and data for this analysis, which we don't show those steps here for the sake of concision in this manuscript, however they are demonstrated in `paper.Rmd` included in the GitHub repository for **hatchR**. After the setup, we can easily map `predict_phenology()` across all putative spawning sites and three spawn dates (September 1-Early Spawning, September 15-Peak Spawning, and September 31-Late Spawning), the results of which are presented in Figure XXX.

@@

With **hatchR** we present a software ecosystem that bridges the analytical gap of predicting developmental phenology for wild fishes and develops a formal framework for applying effective value models from user-provided parameterizations. To do so, the software is bundled in two forms, a fully customizable R package, especially useful for running repetitive analyses and a graphical-user-interface designed for ease of use and tasks that may only run once or a handful of times. Both applications allows users to import their data, run basic data checks, and apply historic model parameterizations for salmonids or create their own species- or population specific parameterization. Additionally, we provide substantial documentation on the **hatchR** website (<https://bmait101.github.io/hatchR/>) to walk users through basic to advanced applications of programmatic platform.

In the application of using effective value models **hatchR** and the user make some key assumptions. Particularly, numerous studies have indicated that stressful environmental conditions can cue fish to prematurely hatch or emerge from their environment. These include water quality like dissolved oxygen, pH, or salinity, pathogens, and even mechanical agitation (reviewed in @quinn2018 and @cowan2024). Moreover, while the **hatchR** provides point estimates for developmental phenology, spawning and developmental within populations of fishes both occur as distributions as opposed single events @mason1976 and we encourage users to either predict phenology using early, peak, and late frameworks (*e.g.*, 5th, 50th, and 95th quantiles) or with real or modeled distributions. Another factor that could bias estimates from these models is that temperatures used from sensors don't reflect the geomoprhic properties in the environment where eggs are developing such that they may be too cold or warm @geist2002.

@@

The models presented in **hatchR** can be further customized in multiple ways beyond the use-cases provided above. For instance, while our models are fit to predict hatch or emergence timing, they could be used to predict other developmental stages prior to the initation of external feeding like eye-up or other embryological developmental stages (@velsen1980) or **Some example from a non-salmonid, maybe initiation of pelagic-larval or riverine current dispersal**. Additionally, while `fit_model()` relies on non-linear modeling platform, `predict_phenology()` only requires users to pass a model expression, so if users preferred a different model fit than the non-linear option provided (as long as it integrates daily temperature), they could pass a different expression into `predict_phenology()` for additional customization. Finally, while **hatchR** was designed specifically for fishes, we expect other poikilothermic organisms such as reptiles, amphibians, or invertebrates would develop in accordance with the power law and these developmental models could be extended to other organisms beyond fishes.

>>>>

Morgan M. Sparks^1,4^, Eli A. Felts^2^, Allison G. Swartz^3^, Bryan M. Maitland^1^

@@

Understanding the timing of key life history events is necessary for managing and conserving populations. Historically, models to predict hatch and emergence timing for fishes were difficult to employ in wild settings because average incubation temperature was needed as the primary parameter in predictive models. However, recent improvements to these techniques reworked models such that they could be applied in wild environments as long as users had data for when adult fish spawned and a record of average daily temperature over the course of development. Despite these improvements, their application remains limited due to few parameterizations for varying species, being largely limited to salmonids. Here we present **hatchR**, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses. **hatchR** allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package. **hatchR** is distributed in two forms—an open source R package for maximum customization and as an HTML graphical-user-interface web application for individuals not familiar with scripting languages. To demonstrate potential uses, we present two case studies as likely applications for this software. **hatchR** promises to open many exciting avenues in research and management of fishes during their early life history

@@

The resulting effective value approach has now been widely applied in Salmonids for which parameterizations from aquaculture were readily available—for example, Pacific Salmon (*Oncorhynchus spp.*) models developed by @beacham1990 have been applied to various species and populations [@adelfio2024; @adelfio2019; @kaylor2021] while models developed for Bull Trout (*Salvelinus confluentus*) by @mcphail1979 were extended by @austin2019. Despite growing popularity, applications have been largely limited within Salmonids, presumably because parameterizations for such models already existed due to their wide use in aquaculture and their general popularity as sport and commercial fish.

To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, **hatchR**. Specifically, **hatchR** allows users to input standard raw or summarized temperature data sets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own or published temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.

To widen the user application of these methods, we distribute two user-interfaces for **hatchR**. The first is a R package (@r:ther) distributed via CRAN that allows users the most customizable application for these methods. The R package is especially powerful as it allows users to automate their analyses over multiple variables such as phenology type, multiple spawn dates, or different habitats with varying thermal regimes. These variable approaches are outlined in the package documentation on **hatchR**'s website. Alternatively, we also distribute a Shiny application (@chang2024) in the form of an HTML-based web tool to interact with many of **hatchR**'s functions in a graphical-user-interface. The Shiny app trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of **hatchR**'s R package. Below, we present the basic overview of the software and multiple case studies of how it may be applied.

@@

**hatchR** is meant to primarily be a tool for predicting phenology. In this sense, we limit functionality to these applications and provide minimal data checking and plotting help. This decision is in part driven by the diversity of data types that users may import and the difficulty in addressing all those data types with respect to various data checks. In other words, we expect users to know their data better than we do and to check it accordingly. We do provide two basic data check functions discussed in the Checking Data section. Similarly, we provide limited functionality for plotting results, but provide examples of how to build custom visualizations from output, specifically in R. For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in software of the user's choice. Finally, we provide brief summaries of general applications of **hatchR** below (Figure 1), but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables.

```{r, echo = FALSE, out.height = '75%', fig.align = 'center', fig.cap="hatchR workflow. Data inputs are represented by folders, data processes by filled circles, hatchR functions as plain text rectangles, and decisions choices as italicized text."}
knitr::include_graphics("./workflow.png")
```

@@

Effective value models were created by @sparks2019 to implement developmental models in wild environments for Sockeye Salmon (*O. nerka*). The need for their creation arose because historic models, specifically those in @beacham1990, only considered the average incubation temperature during development and, for wild fishes, average incubation temperature was impossible to estimate because it was unknown when fish hatched even if adult spawn timing was known. To address this, @sparks2019 used the reciprocal of the formulation of model 2 from @beacham1990 and assigned an effective value for every day of development using the daily average temperature.

@@

Water temperature data sets collected for wild environments are often either 1.) already summarized by day (*i.e.*, mean daily temperature) or, 2.) in a raw format from something like a HOBO TidbiT where readings are taken multiple times per day, which can be summarized into a mean daily temperatures. Alternatively, new statistical models like that of @siegel2023 could be similarly implemented.

@@

**hatchR** assumes you've checked for missing records or errors in your data as it *will function with gaps*, so it's important to go through the data checks discussed below, as well as your own validity checks. **hatchR** can use values down to freezing (e.g, 0 °C), which returns extremely small effective values, and time to hatch or emerge could extend to a year or more. In these cases, we suggest users consider how much of that data type is reasonable with their data.

@@

We demonstrate the utility of the summary and check functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year-long data set (`year_sim`, the dimensions of the data and first 6 rows are shown) with temperature readings every thirty minutes.

@@

@@

```{r}
#year_sim data dimensions
dim(year_sim)
#fist 6 rows of year_sim
head(year_sim)
```

First, we recommend checking imported data for any outliers or strange inputs using `plot_check_temp()` (Figure 2).

```{r, echo = TRUE, fig.cap="Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` data set. Users can set custom thresholds for minimum and maximum temperatures (dashed lines)."}
plot_check_temp(data = year_sim,
                dates = date,
                temperature = temp,
                temp_min = 0, # temp_min and max lines are 
                              # user customizable
                temp_max = 25)
```

@@

Users can either use Salmonid models from `model_table` or generate custom models suing `fit_model()` in both R and Shiny deployments of **hatchR**. As discussed, the models in `model_table` are included because parameterizations already existed in the literature, however, these parameterizations are limited to Pacific Salmon and Bull Trout (though see @quinn2018 pg. 183 for numerous other Salmonids). Additionally, we limit available models in `model_table` because they are well vetted models, with experimental ranges spanning 2-17 °C. Custom models we demonstrate later with `fit_model()` have much narrower parameterization ranges and, as such, we choose to keep those models out of `model_table` so as to require users to consider whether the custom models they are parameterizing fit the data over which they are predicting.

To widen the application of the effective value approach, we include a `fit_model()` function, which is species agnostic (as long as development generally follows the power law). The function `fit_model()` uses data in which average incubation temperature (°C) and days to phenological event (as two vectors) are the inputs. The function estimates parameter coefficients for *log~e~a* and *b* using `stats::nls()`

@@

We demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package. We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for fishes beyond the scope of the original effective value approach. These parameterizations are for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*) [@webster1948], Channel Catfish (*Ictalurus punctatus*) [@small2001], and Lake Sturgeon (*Acipenser fulvescens*) [@smith2005]. We show parameterization for just Smallmouth Bass here for concision, but the full code set for all species is available at in the [paper.Rmd](https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper_pdf/paper.Rmd) in the project GitHub.

@@

```{r, echo=FALSE}
###  make temp regime
set.seed(123)
# create random temps and corresponding dates
temps_sim <- sort(rnorm(n =30, mean = 16, sd = 1), decreasing = FALSE)
dates_sim <-  seq(from = ymd("2000-07-01"),
             to = ymd("2000-07-31"), length.out = 30)

@@

# change names so they aren't the same as the vector objects
colnames(data_sim) <- c("temp_sim", "date_sim")
```

```{r, echo = TRUE}

@@

# change °F to °C and hours to days
smallmouth <- smallmouth |> 
  mutate(days = ceiling(hours/24),
         temp_C = (temp_F -32) * (5/9))

# model object for smallmouth bass 
smb_mod <- fit_model(temp = smallmouth$temp_C,
                     days = smallmouth$days,
                     species = "smb",
                     development_type = "hatch")

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted days to hatch for three warmwater species with custom parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days."}
### predict_phenology

@@

To illustrate model selection and phenology prediction we will recreate a small portion of the analysis done by @sparks2019 using the `woody_island` data set included in this package. We will predict both hatch and emergence timing, so we will obtain a model expression for each using `model_select()`, which calls built-in models from `model_table`.

@@

And then look at the returned object to see days to hatch and development period:

```{r, echo = TRUE}
WI_hatch$days_to_develop
WI_hatch$dev.period
```

@@

`WI_hatch$days_to_develop` outputs the predicted days to develop.

`WI_hatch$dev.period` is a 1x2 dataframe with the dates corresponding to when your fish's parent spawned (which you input with `predict_phenology(spawn.date = ...)`) and the date when the fish is predicted to develop.

`WI_hatch$ef_table` is a *n* x 5 tibble (*n* = number of days to hatch or emerge) and the columns are a row index, the date, each day's temperature and effective value, and the cumulative sum of the effective values. The `ef_table` object is meant to serve as the basis for users to make custom figures for their data beyond the functionality we discuss below.

`WI_hatch$model_specs` provides details about the model used for predicting phenology, which varies depending on if the model was generated from `model_select()` or `fit_model()`, but most importantly contains the model expression (*e.g.*, the formula) used for prediction.

@@

```{r, fig.cap="Output of `plot_phenolgy()` function using predicted hatch time from `woody_island` data set."}
plot_phenology(WI_hatch)
```

@@

We demonstrate this first case study using the graphical user interface portion of the **hatchR** ecosystem found at [https://elifelts.shinyapps.io/hatchR_demo/](https://elifelts.shinyapps.io/hatchR/){.uri}. Users will first upload their data with the `Import Data` window, which requires them to select their file on their personal computer, provide the program with the columns corresponding for dates and temperatures, and then provide the format in which dates are coded (*e.g.*, year-month-day or day-month-year). Data used in this case study are from the `crooked_river` data set. It is included as a demo data set in Shiny, but can also be written out from the package or accessed at <https://github.com/bmait101/hatchR/blob/master/extdata/crooked_river.csv/>. Once data is uploaded the program automatically plots the user's data using `plot_check_temp()` in the background and provides them the outputted graphical check. After uploading and checking data, the user switches to the `Model Phenology` window. In this circumstance, we use the preloaded parameterization for bull trout from @austin2019 with the `Existing` button for model selection, which the user selects with the various drop down options in the menu. After the model is selected, the user can choose multiple spawn dates from the interactive calendar provided. We show results for spawning for September 30th as indicated in the example above. Once dates are chosen, a table entry for each spawn date is outputted in the `Phenology Summaries` tab and corresponding plot with data from each spawn date in the `Timeline Plot` tab. Output from predicting phenology and the resulting figure are downloadable from their respective tabs. The process is demonstrated in the included ***supplementary video file???***, but the interface is described more completely on **hatchR**'s Shiny website.

@@

In this example we expect the last fish out of the gravel well before the June 1st date and the manager could allow road work in this area without worrying about sediment disturbance to fish developing in the gravel.

@@

For the second example, we will again use Bull Trout, but demonstrate a much more complex application for the purpose of showing the full flexibility of the programmatic application of **hatchR**. In this scenario we will use data from @isaak2018 (included `idaho` data set), which includes temperature data from 226 sites across the major upper Columbia River headwater watersheds in Idaho. For this approach we winnow putative bull trout spawning sites by filtering for sites with mean August temperature \</= 13 °C in accordance with thresholds from @isaak2015. For the resulting 139 sites we will demonstrate predicting hatch timing in these putative Bull Trout spawning habitats.

@@

We need to setup our models and data for this analysis, which we don't demonstrate here for the sake of concision, however they are demonstrated in `paper.Rmd` included in the GitHub repository for **hatchR**. After the setup, we can easily map `predict_phenology()` across all putative spawning sites and three spawn dates (September 1-Early Spawning, September 15-Peak Spawning, and September 31-Late Spawning), the results of which are presented in Figure 5.

@@

With **hatchR** we present a software ecosystem that bridges the analytical gap of predicting developmental phenology for wild fishes and develops a formal framework for applying effective value models from user-provided parameterizations. To do so, the software is bundled in two forms, a fully customizable R package, especially useful for running repetitive or complex analyses and a graphical-user-interface designed for ease of use and tasks that may only run once or a handful of times. Both applications allows users to import their data, run basic data checks, and apply historic model parameterizations for Salmonids or create their own species- or population specific parameterizations. Additionally, we provide substantial documentation on the **hatchR** website (<https://bmait101.github.io/hatchR/>) to walk users through basic to advanced applications of the two platforms.

In the application of using effective value models **hatchR** and the user make some key assumptions. Particularly, numerous studies have indicated that stressful environmental conditions can cue fish to prematurely hatch or emerge from their environment. These include water quality like dissolved oxygen, pH, or salinity, pathogens, and even mechanical agitation (reviewed in @quinn2018 and @cowan2024). Moreover, while the **hatchR** provides point estimates for developmental phenology, spawning and developmental within populations of fishes both occur as distributions as opposed single events @mason1976 and we encourage users to either predict phenology using early, peak, and late frameworks (*e.g.*, 5th, 50th, and 95th quantiles) or with real or modeled distributions. Another factor that could bias estimates from these models is that temperatures used from sensors do not reflect the geomorphic properties in the environment where eggs are developing such that they may be too cold or warm @geist2002.

@@

The models presented in **hatchR** can be further customized in multiple ways beyond the use-cases provided above. For instance, while our models are fit to predict hatch or emergence timing, they could be used to predict other developmental stages prior to the initiation of external feeding like eye-up or other embryological developmental stages (@velsen1980) or ***Some example from a non-salmonid, maybe initiation of pelagic-larval or riverine current dispersal***. Additionally, while `fit_model()` relies on non-linear modeling platform, `predict_phenology()` only requires users to pass a model expression, so if users preferred a different model fit than the non-linear option provided (as long as it integrates daily temperature), they could pass a different expression into `predict_phenology()` for additional customization. Finally, while **hatchR** was designed specifically for fishes, we expect other poikilothermic organisms such as reptiles, amphibians, or invertebrates would develop in accordance with the power law and these developmental models could be extended to other organisms beyond fishes.

# Conclusion

**hatchR** bridges the analytical gap for the prediction of developmental phenology for fishes in wild environments. It provides basic data checks and summarization, options for existing or custom model parameterization, and simple to complex phenology prediction frameworks. Importantly, it extends the effective value framework developed by @sparks2019 to a generalizable tool that can be applied to any fish species or population as long as the appropriate source data are available, which can easily be collected in aquaculture settings. We present basic applications for how **hatchR** may be used, with additional and more detailed examples provided on the software's [website](https://bmait101.github.io/hatchR/). **hatchR** is designed for applied and basic applications in which users can engage with the resource in a programmatic environment or a point-and-click Shiny app. We expect we have only demonstrated a small suite of the applications **hatchR** may be used for and encourage the user community to creatively build on the framework developed here.
